---
title: "Credit Risk Modelling: Loan Defaults Prediction"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
    theme: lumen
    highlight: kate
---

```{=html}
<style type="text/css">

body, td {
   font-size: 14px;
}
code.r{
  font-size: 13px;
}
pre {
  font-size: 14px
}
body {
  text-align: justify}
}
</style>

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
```

# Introduction
This dataset contains a variety of information customers regarding credit risk default which is the risk that a lender takes the chance that a borrower fails to make required payments of the loan. In credit risk case, loan default is a problem of unbalanced data. To deal with it, we can use logistic regression. 

There are 32,581 records with 12 columns. This project aims to know risk factors affect loan default and also make prediction probability occurrence or non-occurrence of loan default based on existing predictor values. The data set can be discovered in [here](https://www.kaggle.com/datasets/laotse/credit-risk-dataset)

# Data Preparation
##### Read data
```{r message=FALSE,echo=TRUE}
library(readr)
data <- read.csv("credit_risk_dataset.csv")
```

```{r warning = FALSE, message = FALSE, echo=TRUE}
library(DT)
datatable(data,options=list(pageLength=10,scrollX='400px'),filter='top')
```

##### Description data
-   **`person_age`**: age

-   **`person_income`**: annual income 

-   **`person_home_ownership`**: home onwership

-   **`person_emp_length`**: employment length in years

-   **`loan_intent`**: loan intent

-   **`loan_grade`**: loan_grade

-   **`loan_amnt`**: loan_amount

-   **`loan_int_rate`**: interest rate

-   **`loan_status`**: loan status (0 is non default and 1 is default)

-   **`loan_percent_income`**: percent_income

-   **`cb_person_default_on_file`**: historical default

-   **`cb_person_cred_hist_length`**: credit history length


# Exploratory Data Analysis
```{r message=FALSE,echo=TRUE}
library(dplyr)
glimpse(data)
```

## Transform Data
We can see dataset contain several character variables that need to transformed into factor variables.
```{r message=FALSE,echo=TRUE}
# Encode `loan_status` 0 == non default, 1 == default 
   data$loan_status <-  ifelse(data$loan_status == 0,"non default","default")
```

```{r message=FALSE,echo=TRUE}
# Transform categorical data
library(dplyr)
data <- data %>%
  mutate_at(vars(person_home_ownership,loan_intent,loan_grade,loan_status,cb_person_default_on_file),as.factor)
glimpse(data)
```
## Handling missing values
```{r message=FALSE,echo=TRUE}
colSums(is.na(data))
```
There are missing values in `person_emp_length` and `loan_int_rate`. Some scenarios for handling missing values, one of which we can impute missing values with the most frequent value for categorical variables. While continuous variables impute with the median of the column values.

```{r message=FALSE, echo=TRUE}
#`person_emp_length`, NaN values are replaced with the most frequent value (mode)
data$person_emp_length[is.na(data$person_emp_length)] <- as.numeric(names(which.max(table(data$person_emp_length))))
```

```{r message=FALSE, echo=TRUE}
#`loan_int_rate`, NaN values are replaced with the the median of the column values 
data$loan_int_rate[is.na(data$loan_int_rate)] <- median(data$loan_int_rate,na.rm=TRUE)
```
```{r message=FALSE,echo=TRUE}
# Check missing values are replaced
colSums(is.na(data))
```

All variables with different data types have been converted to the desired data type and there are no missing values.

## Outliers
```{r message=FALSE, echo=TRUE}
library(ggplot2)
library(ggpubr)
library(patchwork)
```
```{r message=FALSE, echo=TRUE}
fig_1 <- ggplot(data,aes(x=person_age)) + 
            geom_histogram(aes(y=after_stat(density)),color="skyblue",fill="skyblue") + geom_density(alpha=0.2) + ggtitle("Figure 1: The density of age person")
fig_1
```
<br>

Figure 1 shows the majority of people are 20 to 60 years old. People who have age more than 120 can be indicated outliers and it will be deleted. 

```{r message=FALSE, echo=TRUE}
fig_2 <- ggplot(data,aes(x=person_income)) + 
            geom_histogram(aes(y=after_stat(density)),color="skyblue",fill="skyblue") +
            geom_density(alpha=0.2,fill="skyblue") + ggtitle("Figure 2: The density of annual income person")
fig_2
```
<br>

Figure 2 depicts there is indication possible outliers to people who have income more than 4 millions and it will be dropped.

```{r message=FALSE, echo=TRUE}
fig_3 <- ggplot(data,aes(x=person_emp_length)) +
            geom_histogram(aes(y=after_stat(density)),color="skyblue",fill="skyblue") +
            geom_density(alpha=0.2,fill="skyblue") + ggtitle("Figure 3: The density of employment length")
fig_3
```

<br>

Figure 3 represents there is also indication possible outliers. People have employment length more than 60 years will be removed.

```{r message=FALSE, echo=TRUE}
# Outliers will be removed 
person_age_out <- which(data$person_age > 100) 
person_income_out <- which(data$person_income > 4000000)
emp_length_out <- which(data$person_emp_length > 60)
data_cr <- data[-c(person_age_out,person_income_out,emp_length_out),]
```
## Visualization 
```{r message=FALSE,echo=TRUE}
library(ggcorrplot)
model.matrix(~0+., data=data_cr) %>% 
  cor(use="pairwise.complete.obs") %>% 
  ggcorrplot(show.diag = FALSE, type="lower",tl.cex=5, lab=TRUE,
             lab_size=1.5) + ggtitle("Figure 4: Correlation between variables")
```
<br>
As we can see, some variables have correlaction to `loan_status` such as `person_home_ownership`,`loan_intent`,`loan_grade`,`loan_amnt`, and `loan_int_rate`.

```{r message=FALSE,echo=TRUE}
fig_loan <- ggplot(data_cr,aes(x=loan_status, fill=loan_status)) + geom_bar() + 
               geom_text(stat = "count", aes(label = after_stat(count)), 
               colour = "white", size = 3.5, position=position_stack(vjust=0.5)) + 
               ggtitle("Figure 5: Loan Status")
fig_loan
```
<br>

The count of people with default loan is less than those with non-default status. 

```{r message=FALSE,echo=TRUE}
# Relation between person_home_ownership with loan status
fig_ph <- ggplot(data_cr,aes(x=person_home_ownership, fill=loan_status)) + geom_bar() + 
             geom_text(stat = "count", aes(label = after_stat(count)), 
             colour = "white", size = 2.5, position=position_stack(vjust=0.5)) + 
             ggtitle("Figure 6 : Person home ownership by loan status")

# Relation between loan grade with loan status
fig_grade <- ggplot(data_cr,aes(x=loan_grade, fill=loan_status)) + geom_bar() + 
                geom_text(stat = "count", aes(label = after_stat(count)), 
                colour = "white", size = 2.5, position=position_stack(vjust=0.5)) +
                ggtitle("Figure 7: Loan grade by loan status")

fig_mix_1 <- fig_ph / fig_grade
fig_mix_1
```
<br>

Figure 6 displays mostly people with default and non-default loan have mortgage and rent for home ownership. In figure 7, the number of people with non-default loan decrease as the loan grade increasing. It occurs the increasing and decreasing in the number of people with default loan as the loan grade enhancing.  

```{r message=FALSE,echo=TRUE}
# Relation between loan intent with loan status
fig_int <- ggplot(data_cr,aes(x=loan_intent, fill=loan_status)) + geom_bar() + 
              geom_text(stat = "count", aes(label = after_stat(count)), 
              colour = "white", size = 2.5, position=position_stack(vjust=0.5)) + 
              theme(axis.text.x = element_text(angle = 90)) +
              ggtitle("Figure 8: Loan intent by loan status")
fig_int
```
<br> 

In figure 8, people with non-default more tend to take loan for education  necessary and people with default mostly loan used for medical. 


```{r message=FALSE,echo=TRUE}
# Relation between interest rate with loan status
fig_ir <- ggplot(data_cr,aes(x=loan_status, y=loan_int_rate, fill=loan_status)) + geom_boxplot() + 
             ggtitle("Figure 9: Interest rate by loan status") + coord_flip()

# Relation between loan amount with loan status
fig_am <- ggplot(data_cr,aes(x=loan_status, y=loan_amnt, fill=loan_status)) + geom_boxplot() +
             ggtitle("Figure 10: Loan amount by loan status") + coord_flip()

fig_mix_2 <- fig_ir / fig_am
fig_mix_2
```

<br>

The boxplot in figure 9 above appears to indicate that people with default have a higher interest rate than people with non-default. Figure 10 depicts that people with non-default have less than loan amount than people with default. 

# Modelling {.tabset}
## Splitting dataset
We can split dataset into the training set used to build model and testing set used to examine the model has been obtained. We can divide 80% for training set and 20% for testing set. 

```{r message=FALSE, echo=TRUE}
library(dplyr)
set.seed(123)
split <- sample(nrow(data_cr),size=nrow(data_cr)*0.8)
train <- data_cr[split,]
head(train)
test <- data_cr[-split,]
head(test)
```

## Building model
```{r message=FALSE, echo=TRUE}
model_loan <- glm(formula = loan_status  ~ . ,family="binomial", data = train)
summary(model_loan)
```

##### Simultaneous Parameter Significance Test
This section explains testing to check coefficient significance using hypothesis null is there is no predictor variables affect response variable (loan status). Alternative hypothesis is at least one predictor variable influences loan status.

```{r message=FALSE, echo=TRUE}
library(pscl)
pR2(model_loan,c("McFadden"))
cat('chi square value:',qchisq(0.95,11))
```

Based on result performs model loan has $G^{2}$ value more than chi-square table value with significance level 0.05. Therefore, decision to reject hypothesis null means at least one predictor variable affects significantly to status loan. 

##### Partial Parameter Significance Test
We do testing to know significance each predictor variables to response variable. Hypothesis null is the ***i-th*** predictor variables doesn't have significant influencing to response variable (loan status). While, alternative hypothesis is at least one ***i-th*** predictor variable has significant influencing to loan status.

```{r message=FALSE, echo=TRUE}
summary(model_loan)
```

Model loan represents that 8 out of the 11 predictors are significantly associated to loan status. Variables exclude: `person_age`, `cb_person_default_on_file` for Y category, and `cb_person_cred_hist_length`. 

##### Final Model
```{r message=FALSE, echo=TRUE}
#Rebuilding model using significant predictor variables to loan status
model_loan_final <- glm(formula = loan_status ~  person_income + person_home_ownership + person_emp_length + loan_intent + loan_grade +
                         loan_amnt + loan_int_rate + loan_percent_income,
                       family="binomial", data = train)
summary(model_loan_final)
```
```{r message=FALSE, echo=TRUE}
# Simultaneous Parameter Significance Test
library(pscl)
pR2(model_loan_final,c("McFadden"))
cat('chi-square value:',qchisq(0.95,8))
```
```{r message=FALSE, echo=TRUE}
# VIF values
library(car)
vif(model_loan_final)
```
The predictor variables in model have a VIF value less 5, we can assume that multicollinearity is not an issue in the final model.

```{r message=FALSE, echo=TRUE}
# Partial Parameter Significance Test
summary(model_loan_final)
```
Model final loan loads eight predictor variables can be said as best model for further analysis due to meet parameter significance test simultaneously and partially.

# Model Evaluation {.tabset}
## Odd Ratio
Calculation odd ratio is used to obtain model interpretation.

```{r message=FALSE, echo=TRUE}
coef_loan <- coef(model_loan_final)
odd_model_loan <- cbind(coef_loan, OR = exp(coef_loan), SK= exp(confint(model_loan_final)))
odd_model_loan
```
## Prediction
Performance model can be evaluated with make prediction using testing set. 
```{r message=FALSE, echo=TRUE}
prob_loan <- predict(model_loan_final, type= "response", newdata = test)
pred_loan <- ifelse(prob_loan > 0.5, "non default", "default")
test_loan <- data.frame(pred_loan,test$loan_status)
head(test_loan)
```

```{r message=FALSE, echo=TRUE}
loan <- table(pred_values = pred_loan, actual_values = test$loan_status)
loan
```
Confusion Matrix output represents for predict loan status category default can predict correctly 818 out of 1473 while for category non default can predict correctly 4814 out of 5042. 

## Model diagnostics
```{r message=FALSE, echo=TRUE}
#number of true positives/(number of true positives + number of false negatives)
sensitivity <- loan[2,2]/sum(loan[2,])*100 

#number of true negatives/(number of true negatives + number of false positives)
specificity <- loan[1,1]/sum(loan[1,])*100 

#number of true negatives + number of true positives/ (number of true positives + number of false positives + number of true negatives + number of false negatives)
accuracy <- (loan[1,1]+loan[2,2])/sum(loan)*100

fprate <- loan[2,1]/(loan[2,1]+loan[1,1])*100
AUC <- (100+sensitivity-fprate)/2 
performance <- data.frame(sensitivity,specificity,accuracy,AUC)
performance
```
Sensitivity or True Positive Rate (TPR) is percentage of actual positives which are correctly identified. In model, TPR value shows that we detect correctly default loan status 88.02%. 

While, Specificity or True Negative Rate (TNR) percentage of actual negatives will test negative. It can be shown FPR value that we detect correctly non default loan status 78.20%. 

The classification model has a good prediction capability to default loan status and has a good enough prediction capability to non default loan status. 

Accuracy 86.45% is good for correctly predicted using binary logistic regression model. 

